DRAFT_Cleaning

### 2025-04-29 Draft_Cleaning.Rmd
reran some draft cleaning stuff

```{r setup, include=FALSE}
library(readr)
library(tidyverse)
library(lubridate)
library(caret)
library(ggfortify)
library(ggformula)
library(DiagrammeR)
library(gridExtra)

@@

### Read in the data
```{r}
dredge_data<-read_csv("DredgeData.csv", show_col_types = FALSE)
survey_data<-read_csv("SurveyData.csv", show_col_types = FALSE)
gage_data<-read_csv("UMR_IWW_1999_2021.csv", show_col_types = FALSE)
```

@@

### Cleaning the data
Now lets clean the survey data. I only want conditional survey records between
1999 to 2000, these are going to act as my background data. I want to combine
these records with my dredge data. I'm also going to filter by river, remove
pools that are not of interest, and create some 
columns to support my join. Dates are also formatted as date objects and some
columns are renamed to make it easier to join my surveys to my dredge data.
Finally, select the columns we need for our analysis to join to our dredging 
data. 
```{r}
survey_data<- survey_data |>
  filter(SURVEYTYPE == "conditional",
         SURVEY_YEAR >= 1999 & SURVEY_YEAR <= 2022,
          RIVER == "Mississippi_River" | RIVER == "Illinois_Waterway",
         !POOL %in% c("LP","CS"))|>
  mutate(VOLUMEDREDGED = 0,
         DREDGINGPURPOSE = "None",
         DREDGE_TYPE = "None",
         LABOR_TYPE = LABOR_SOURCE,
         EXECYEAR = SURVEY_YEAR,
         FEATURENAME = SDSFEATURENAME,
         DATE_START = as.character(DATE_START),
         DATE_START = as_date(DATE_START)
         )|>
  select(EXECYEAR,DATE_START,DREDGINGPURPOSE,VOLUMEDREDGED, RIVER,POOL, DREDGE_TYPE,
         UP_RIV_MIL,DN_RIV_MIL)
```

Lets clean the dredge data, we only want records that are channel dredging events
and since the dredge records only go back to 1999, we just need to filter out data beyond 2022.
```{r}
dredge_data <- dredge_data |>
   filter(
          DREDGINGPURPOSE != "harbor",
           EXECYEAR <= 2022)|>
  mutate(
     DATE_START = as_date(mdy_hm(DATE_START)))|>
  select(EXECYEAR,DREDGINGPURPOSE, VOLUMEDREDGED, RIVER,POOL, DREDGE_TYPE,
         DATE_START, UP_RIV_MIL,DN_RIV_MIL)
```

@@

combined_data <-combined_data |>
     filter(DATE_START >= as.Date("1999-04-01") &
         DATE_START <= as.Date("2021-09-30")) |>
          mutate( DREDGINGPURPOSE = case_when(DREDGINGPURPOSE == "None" ~ "No Dredging",
                                      DREDGINGPURPOSE == "channel" ~ "Dredged"))

@@

gage_survey_filt <- gage_survey |>
  filter(complete.cases(gage_survey))
```

@@

```
Now on to analysis!

@@

fit_pca = train(VOLUMEDREDGED ~ .,
            data = gage_survey_PCA,
            method = "glm",
            preProc = c("pca"),
            trControl = ctrl)
fit_pca

@@

ctrl = trainControl(method = "cv", number = 5) 
fit_dredge = train(VOLUMEDREDGED ~ .,
                 data = gage_xgb,
                 method = "xgbTree",
               tuneGrid = expand.grid(nrounds = c (15,25,50,100), 
                                              max_depth = 1:10,
                                        eta =c(0.3), gamma = 0,
                                        colsample_bytree = 1,
                                        min_child_weight = 1, subsample = 1),
                 verbosity = 0,
                 trControl = ctrl,
                 na.action = na.exclude)

@@

  ctrl = trainControl(method = "cv", number = 5)
  fit_dredge_DCV = train(VOLUMEDREDGED ~ .,
                       data = gage_xgb[in_train, ],
                       method = "xgbTree",
                       tuneGrid = expand.grid(nrounds = c (15,25,50,100), 
                                              max_depth = 1:10,
                                        eta =c(0.3), gamma = 0,
                                        colsample_bytree = 1,
                                        min_child_weight = 1, subsample = 1),
                       verbosity = 0,
                       trControl = ctrl,
                       na.action = na.exclude)

@@

combined_data <-combined_data |>
     filter(DATE_START >= as.Date("1999-04-01") &
         DATE_START <= as.Date("2021-09-30"))

@@

gage_IWW <- gage_data |>
  select(DATE_START,BEAI2,HAVI2,
         HNYI2,IL03, IL03_Tail, IL04, IL04_Tail, IL05, IL05_Tail,IL06, IL06_Tail,
         IL07, IL07_Tail, IL08, IL08_Tail,KNGI2, MORI2, PIAI2)

gage_Miss <- gage_data |>
  select(DATE_START,BRLI4,CMMI4, DBQI4,
         FAII4, KHBI2,MI11,MI11_Tail,MI12,MI12_Tail,MI13,MI13_Tail,MI14,MI14_Tail,
         MI15,MI15_Tail,MI16,MI16_Tail,MI17,MI17_Tail,MI18,MI18_Tail,MI19,MI19_Tail,
         MI20,MI20_Tail,MI21,MI21_Tail,MI22,MI22_Tail, MUSI4,UINI2)

Illinois_River <- gage_survey |>
  select(EXECYEAR,DATE_START,DREDGINGPURPOSE,VOLUMEDREDGED,RIVER,
         POOL, DREDGE_TYPE,UP_RIV_MIL, DN_RIV_MIL,lead_7,lead_14,BEAI2,HAVI2,
         HNYI2,IL03, IL03_Tail, IL04, IL04_Tail, IL05, IL05_Tail,IL06, IL06_Tail,
         IL07, IL07_Tail, IL08, IL08_Tail,KNGI2, MORI2, PIAI2) |>
  filter(RIVER == "Illinois_Waterway")

Miss_River <- gage_survey |>
  select(EXECYEAR,DATE_START,DREDGINGPURPOSE,VOLUMEDREDGED,RIVER,
         POOL, DREDGE_TYPE,UP_RIV_MIL, DN_RIV_MIL,lead_7,lead_14,BRLI4,CMMI4, DBQI4,
         FAII4, KHBI2,MI11,MI11_Tail,MI12,MI12_Tail,MI13,MI13_Tail,MI14,MI14_Tail,
         MI15,MI15_Tail,MI16,MI16_Tail,MI17,MI17_Tail,MI18,MI18_Tail,MI19,MI19_Tail,
         MI20,MI20_Tail,MI21,MI21_Tail,MI22,MI22_Tail, MUSI4,UINI2) |>
  filter(RIVER == "Mississippi_River")

@@

ctrl = trainControl(method = "cv", number = 10) 
Miss_xgb_fit = train(VOLUMEDREDGED ~ .,
                 data = Miss_xgb,
                 method = "xgbTree",
               tuneGrid = expand.grid(nrounds = c(15,25,50,100), 
                                              max_depth = 1:10,
                                        eta =c(0.3), gamma = 0,
                                        colsample_bytree = 1,
                                        min_child_weight = 1, subsample = 1),
                 verbosity = 0,
                 trControl = ctrl,
                 na.action = na.exclude)

@@

ctrl = trainControl(method = "cv", number = 10) 
IWW_xgb_fit = train(VOLUMEDREDGED ~ .,
                 data = IWW_xgb,
                 method = "xgbTree",
               tuneGrid = expand.grid(nrounds = c(15,25,50,100), 
                                              max_depth = 1:10,
                                        eta =c(0.3), gamma = 0,
                                        colsample_bytree = 1,
                                        min_child_weight = 1, subsample = 1),
                 verbosity = 0,
                 trControl = ctrl,
                 na.action = na.exclude)

@@

import_matrix = xgb.importance(model = IWW_xgb_fit$finalModel)
import_matrix
xgb.plot.importance(import_matrix)
```

>>>>

```{r setup, include=FALSE}
library(readr)
library(tidyverse)
library(lubridate)
library(caret)
library(ggfortify)
library(ggformula)
library(DiagrammeR)
library(gridExtra)
library(keras)
library(tensorflow)

@@

### Read in the data
```{r}
dredge_data<-read_csv("DredgeData.csv", show_col_types = FALSE)
survey_data<-read_csv("SurveyData.csv", show_col_types = FALSE)
gage_data<-read_csv("UMR_IWW_1999_2024.csv", show_col_types = FALSE)
```

@@

### Cleaning the data
Now lets clean the survey data. I only want conditional survey records between
1999 to 2022, these are going to act as my background data. I want to combine
these records with my dredge data. I'm also going to filter by river, remove
pools that are not of interest, and create some 
columns to support my join. Dates are also formatted as date objects and some
columns are renamed to make it easier to join my surveys to my dredge data.
Finally, select the columns we need for our analysis to join to our dredging 
data. 
```{r}
survey_data<- survey_data |>
  filter(SURVEYTYPE == "conditional",
         SURVEY_YEAR >= 1999 & SURVEY_YEAR <= 2024,
          RIVER == "Mississippi_River" | RIVER == "Illinois_Waterway",
         !POOL %in% c("LP","CS"))|>
  mutate(VOLUMEDREDGED = 0,
         DREDGINGPURPOSE = "None",
         DREDGE_TYPE = "None",
         LABOR_TYPE = LABOR_SOURCE,
         EXECYEAR = SURVEY_YEAR,
         FEATURENAME = SDSFEATURENAME,
         DATE_START = as.character(DATE_START),
         DATE_START = as_date(DATE_START)
         )|>
  select(EXECYEAR,DATE_START,DREDGINGPURPOSE,VOLUMEDREDGED, RIVER,POOL, DREDGE_TYPE,
         UP_RIV_MIL,DN_RIV_MIL)
```

Lets clean the dredge data, we only want records that are channel dredging events
and since the dredge records only go back to 1999, we just need to filter out data beyond 2022.
```{r}
dredge_data <- dredge_data |>
   filter(
          DREDGINGPURPOSE != "harbor",
           EXECYEAR <= 2024)|>
  mutate(
     DATE_START = as_date(mdy_hm(DATE_START)))|>
  select(EXECYEAR,DREDGINGPURPOSE, VOLUMEDREDGED, RIVER,POOL, DREDGE_TYPE,
         DATE_START, UP_RIV_MIL,DN_RIV_MIL)
```

@@

combined_data <-combined_data |>
     filter(DATE_START >= as.Date("1999-04-01") &
         DATE_START <= as.Date("2024-12-31")) |>
          mutate( DREDGINGPURPOSE = case_when(DREDGINGPURPOSE == "None" ~ "No Dredging",
                                      DREDGINGPURPOSE == "channel" ~ "Dredged"))

@@

gage_survey_filt <- gage_survey |>
  filter(complete.cases(gage_survey))

write_csv(gage_survey_filt,"Gage_Survey_Dredge.csv")
```

@@

Lets make some plots of the data!
```{r}
gages_pivoted <- gage_data|>
    pivot_longer(cols = -DATE_START, 
               names_to = "gage", 
               values_to = "value")

gage_survey_cleaned |>
  ggplot(aes(y=VOLUMEDREDGED, x=DATE_START))+
  geom_line()

gages_pivoted |>
  ggplot(aes(y = value, x = DATE_START, color = gage, group = gage)) +
  geom_line(size = 1) +  # Adjust line thickness
  scale_color_viridis_d() +  # A color scale that's visually appealing
  labs(
    title = "Time Series of Gage Values",
    x = "Date",
    y = "Value",
    color = "Gage"
  ) +  # Adding titles and labels
  theme_minimal() +  # Use a minimal theme
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis labels for better readability
    axis.title = element_text(size = 12),  # Increase title font size
    legend.position = "right",  # Position the legend on the right
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10)
  )
```

```{r}
gage_survey |>
  filter(RIVER == "Mississippi_River")|>
  ggplot(aes(y=VOLUMEDREDGED, x=DATE_START,color = POOL))+
  geom_point()+
  labs(title = "Miss Dredging")

gage_survey |>
  filter(RIVER == "Illinois_Waterway")|>
  ggplot(aes(y=VOLUMEDREDGED, x=DATE_START, color = POOL))+
  geom_point()+
  labs(title = "IWW Dredging")

Now on to analysis!

@@

fit_pca = caret::train(VOLUMEDREDGED ~ .,
            data = gage_survey_PCA,
            method = "glm",
            preProc = c("pca"),
            trControl = ctrl)
fit_pca

@@

ctrl = trainControl(method = "cv", number = 5) 
fit_dredge = caret::train(VOLUMEDREDGED ~ .,
                 data = gage_xgb,
                 method = "xgbTree",
               tuneGrid = expand.grid(nrounds = c (15,25,50,100), 
                                              max_depth = 1:10,
                                        eta =c(0.3), gamma = 0,
                                        colsample_bytree = 1,
                                        min_child_weight = 1, subsample = 1),
                 verbosity = 0,
                 trControl = ctrl,
                 na.action = na.exclude)

@@

  ctrl = trainControl(method = "cv", number = 5)
  fit_dredge_DCV = caret::train(VOLUMEDREDGED ~ .,
                       data = gage_xgb[in_train, ],
                       method = "xgbTree",
                       tuneGrid = expand.grid(nrounds = c (15,25,50,100), 
                                              max_depth = 1:10,
                                        eta =c(0.3), gamma = 0,
                                        colsample_bytree = 1,
                                        min_child_weight = 1, subsample = 1),
                       verbosity = 0,
                       trControl = ctrl,
                       na.action = na.exclude)

@@

combined_data <-combined_data |>
     filter(DATE_START >= as.Date("1999-04-01") &
         DATE_START <= as.Date("2024-12-31"))

@@

gage_IWW <- gage_data |>
  select(DATE_START,BEAI2,HAVI2,
         HNYI2,IL03, IL03_Tail, IL04, IL04_Tail, IL05, IL05_Tail,IL06, IL06_Tail,
         IL07, IL07_Tail, IL08, IL08_Tail,KNGI2, MORI2, PIAI2,"05552500","05527500",
         "05585000","05583000","05570000","05555300")

gage_Miss <- gage_data |>
  select(DATE_START,BRLI4,CMMI4, DBQI4,
         FAII4, KHBI2,MI11,MI11_Tail,MI12,MI12_Tail,MI13,MI13_Tail,MI14,MI14_Tail,
         MI15,MI15_Tail,MI16,MI16_Tail,MI17,MI17_Tail,MI18,MI18_Tail,MI19,MI19_Tail,
         MI20,MI20_Tail,MI21,MI21_Tail,MI22,MI22_Tail, MUSI4,UINI2,"05446500")

Illinois_River <- gage_survey |>
  select(EXECYEAR,DATE_START,DREDGINGPURPOSE,VOLUMEDREDGED,RIVER,
         POOL, DREDGE_TYPE,UP_RIV_MIL, DN_RIV_MIL,lead_7,lead_14,BEAI2,HAVI2,
         HNYI2,IL03, IL03_Tail, IL04, IL04_Tail, IL05, IL05_Tail,IL06, IL06_Tail,
         IL07, IL07_Tail, IL08, IL08_Tail,KNGI2, MORI2, PIAI2,"05552500","05527500",
         "05585000","05583000","05570000","05555300") |>
  filter(RIVER == "Illinois_Waterway")

Miss_River <- gage_survey |>
  select(EXECYEAR,DATE_START,DREDGINGPURPOSE,VOLUMEDREDGED,RIVER,
         POOL, DREDGE_TYPE,UP_RIV_MIL, DN_RIV_MIL,lead_7,lead_14,BRLI4,CMMI4, DBQI4,
         FAII4, KHBI2,MI11,MI11_Tail,MI12,MI12_Tail,MI13,MI13_Tail,MI14,MI14_Tail,
         MI15,MI15_Tail,MI16,MI16_Tail,MI17,MI17_Tail,MI18,MI18_Tail,MI19,MI19_Tail,
         MI20,MI20_Tail,MI21,MI21_Tail,MI22,MI22_Tail, MUSI4,UINI2,"05446500") |>
  filter(RIVER == "Mississippi_River")

@@

ctrl = trainControl(method = "cv", number = 10) 
Miss_xgb_fit = caret::train(VOLUMEDREDGED ~ .,
                 data = Miss_xgb,
                 method = "xgbTree",
               tuneGrid = expand.grid(nrounds = c(15,25,50,100), 
                                              max_depth = 1:10,
                                        eta =c(0.3), gamma = 0,
                                        colsample_bytree = 1,
                                        min_child_weight = 1, subsample = 1),
                 verbosity = 0,
                 trControl = ctrl,
                 na.action = na.exclude)

@@

ctrl = trainControl(method = "cv", number = 10) 
IWW_xgb_fit = caret::train(VOLUMEDREDGED ~ .,
                 data = IWW_xgb,
                 method = "xgbTree",
               tuneGrid = expand.grid(nrounds = c(15,25,50,100), 
                                              max_depth = 1:10,
                                        eta =c(0.3), gamma = 0,
                                        colsample_bytree = 1,
                                        min_child_weight = 1, subsample = 1),
                 verbosity = 0,
                 trControl = ctrl,
                 na.action = na.exclude)

@@

import_matrix = xgb.importance(model = IWW_xgb_fit$finalModel)
import_matrix
xgb.plot.importance(import_matrix)
```
###LSTM
```{r}
library(tensorflow)
library(keras)
model_7 <- keras_model_sequential() |>
  layer_lstm(units = 50, return_sequences = TRUE, input_shape = c(7, ncol(gage_xgb) - 1))  |>
  layer_dropout(rate = 0.2)|>
  layer_lstm(units = 50, return_sequences = TRUE) |>
  layer_dropout(rate = 0.2) |>
  layer_lstm(units = 50) |>
  layer_dropout(rate = 0.2) |>
  layer_dense(units = 1)

model_7 %>% compile(
  loss = 'mean_squared_error',
  optimizer = 'adam'
)

# Function to create sequences based on window size
create_sequences <- function(data, target_column, sequence_length) {
  X <- list()
  y <- list()

  for (i in (sequence_length + 1):nrow(data)) {
    X[[i - sequence_length]] <- as.matrix(data[(i - sequence_length):(i - 1), -target_column])  # Features excluding target
    y[[i - sequence_length]] <- data[i, target_column]  # Target column
  }

  X <- array(unlist(X), dim = c(length(X), sequence_length, ncol(data) - 1))  # Reshape to 3D array
  y <- unlist(y)

  return(list(X = X, y = y))
}

# Example: Create sequences for 7-day and 14-day windows
window_size_7 <- 7
window_size_14 <- 14

# Load necessary libraries
library(recipes)

# Assume gage_xgb is your original dataset
gage_xgb_scaled_recipe <- recipe(~ ., data = gage_xgb) %>%
  # Define step to scale all numeric predictors
  step_scale(all_numeric_predictors()) %>%
  # You can also center them if needed
  step_center(all_numeric_predictors())

# Prep the recipe
gage_xgb_prepped <- prep(gage_xgb_scaled_recipe, training = gage_xgb)

# Apply the scaling to the data
gage_xgb_scaled <- bake(gage_xgb_prepped, new_data = gage_xgb)

# View scaled data
head(gage_xgb_scaled)

# Assume `gage_xgb_scaled` is your scaled dataset and the target column is 'VOLUMEDREDGED'
data_windows_7 <- create_sequences(gage_xgb_scaled, target_column = which(names(gage_xgb) == "VOLUMEDREDGED"), sequence_length = window_size_7)
data_windows_14 <- create_sequences(gage_xgb_scaled, target_column = which(names(gage_xgb) == "VOLUMEDREDGED"), sequence_length = window_size_14)

# X_train and y_train for 7-day window
X_train_7 <- data_windows_7$X
y_train_7 <- data_windows_7$y

# X_train and y_train for 14-day window
X_train_14 <- data_windows_14$X
y_train_14 <- data_windows_14$y

# Define the train-test split
train_size_7 <- floor(0.8 * length(y_train_7))  # 80% for training
train_size_14 <- floor(0.8 * length(y_train_14))  # 80% for training

# Training and Testing Data for 7-day window
X_train_7_final <- X_train_7[1:train_size_7,,]
y_train_7_final <- y_train_7[1:train_size_7]
X_test_7_final <- X_train_7[(train_size_7+1):length(y_train_7),,]
y_test_7_final <- y_train_7[(train_size_7+1):length(y_train_7)]

# Training and Testing Data for 14-day window
X_train_14_final <- X_train_14[1:train_size_14,,]
y_train_14_final <- y_train_14[1:train_size_14]
X_test_14_final <- X_train_14[(train_size_14+1):length(y_train_14),,]
y_test_14_final <- y_train_14[(train_size_14+1):length(y_train_14)]

# Train the model for the 7-day window
history_7 <- model_7 %>% fit(
  X_train_7_final, 
  y_train_7_final,
  epochs = 20, 
  batch_size = 32, 
  validation_data = list(X_test_7_final, y_test_7_final)
)

# Evaluate model on 7-day window
score_7 <- model_7 %>% evaluate(X_test_7_final, y_test_7_final)
cat('Test loss for 7-day window:', score_7, "\n")

predictions_7 <- model_7 %>% predict(X_test_7_final)
```
